Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2023-07-26 analysis/paper/paper.tex
prepare supp mat for submission

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{lineno}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy1,; Jørgen B. Hasselstrøm2; Lambert K. Sørensen2; Maia Casna1; Menno Hoogland1; Historisch Genootschap Beemster3; Amanda G. Henry1},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}
\author{Bjørn Peare Bartholdy\textsuperscript{1,*} \and Jørgen B. Hasselstrøm\textsuperscript{2} \and Lambert K. Sørensen\textsuperscript{2} \and Maia Casna\textsuperscript{1} \and Menno Hoogland\textsuperscript{1} \and Historisch Genootschap Beemster\textsuperscript{3} \and Amanda G. Henry\textsuperscript{1}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 60\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, children and women.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

All statistical analysis was conducted in R version 4.3.0 (2023-04-21),
Already Tomorrow, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Hadley Wickham et al., 2019) and visualisations
were created using \textbf{ggplot2} (H. Wickham, 2016). Polychoric
correlations were calculated with the \textbf{psych} package (Revelle,
2022).

@@

\hypertarget{tbl-compound-detect}{}
\begin{longtable}[]{@{}lllr@{}}
\caption{\label{tbl-compound-detect}Target compound including whether it
was detected (TRUE) or not (FALSE) in each batch, as well as the lower
limit of quantitation (LLOQ) in ng. CBD = cannabidiol; CBN = cannabinol;
THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A;
THCVA = tetrahydrocannabivarin acid.}\tabularnewline
\toprule()
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule()
\endfirsthead
\toprule()
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule()
\endhead
CBD & TRUE & FALSE & 0.050 \\
CBN & TRUE & FALSE & 0.050 \\
Caffeine & TRUE & TRUE & 0.050 \\
Cocaine & TRUE & FALSE & 0.025 \\
Cotinine & TRUE & TRUE & 0.050 \\
Nicotine & TRUE & TRUE & 0.100 \\
Salicylic acid & TRUE & TRUE & 0.500 \\
THC & TRUE & FALSE & 0.100 \\
THCA-A & TRUE & FALSE & 0.025 \\
THCVA & TRUE & FALSE & 0.010 \\
Theophylline & TRUE & TRUE & 0.010 \\
\bottomrule()
\end{longtable}

@@

\hypertarget{tbl-pearson}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1566}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1566}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}@{}}
\caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
dichotomous skeletal lesions and compound concentrations (ng/mg) from
the second batch. Correlations between pairs of dichotomous variables
are removed due to incompatibility with a Pearson correlation. OA =
osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD
= degenerative disc disease; CO = cribra orbitalia; CMS = chronic
maxillary sinusitis; SA = salicylic acid; PN = pipe
notches.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule()
\endhead
OA & -0.19 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
VOP & -0.061 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
SN & -0.22 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
DDD & 0.032 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 & -0.0086 \\
CO & 0.14 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
CMS & -0.18 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
Caries & & -0.13 & -0.27 & -0.19 & -0.037 & -0.16 & 0.079 & -0.16 \\
Nicotine & & & -0.21 & 0.01 & -0.014 & 0.43 & 0.14 & 0.98 \\
SA & & & & 0.14 & 0.37 & 0.038 & 0.17 & -0.17 \\
Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.031 \\
PN & & & & & & -0.16 & 0.18 & -0.0068 \\
Theophylline & & & & & & & 0.51 & 0.36 \\
Caffeine & & & & & & & & 0.078 \\
\bottomrule()
\end{longtable}

@@

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.851). Moderate correlations were
found between OA and DDD (0.484), VOP and periodontitis (0.491), SN and
cotinine (0.558), DDD and calculus (-0.421), CMS and caffeine (0.528),
caries and periodontitis (0.486), caries and theophylline (-0.491),
periodontitis and caries (0.486), periodontitis and caffeine (0.515),
nicotine and CMS (0.496), calculus and caries (0.502), age-at-death and
theophylline (-0.447), theophylline and age-at-death (-0.447), caffeine
and periodontitis (0.515), cotinine and CMS (0.425). Remaining
correlations were weak or absent (Figure~\ref{fig-polycorr}).
Correlations with age will be depressed because age was largely
controlled for in the sample selection.

@@

Because we targeted individuals with moderate-to-large calculus
deposits, it is likely a biased sample. The presence of calculus may
increase the risk of premature death (Yaussy \& DeWitte, 2019), and
periodontal disease (which may or may not be associated with dental
calculus build-up) is a risk-factor for respiratory diseases, if
periodontal and respiratory pathogens enter the bloodstream (Azarpazhooh
\& Leake, 2006; Scannapieco, 1999; Scannapieco \& Ho, 2001). In our
sample, the percentage of chronic maxillary sinusitis (37.0\%) is lower
than in another (more representative) male sample (44.1\%) (Casna et
al., 2021), and the caries percentage is similarly lower in our sample
(12.7\%) than a more representative sample (22.9\%) (Lemmers et al.,
2013).\\
We used the presence/absence of a pipe notch and concurrent detection of
tobacco as a crude estimate of the accuracy of the method, which we
found to be around 59.3\%. This is a very rough estimate, as the
presence of a pipe notch is likely not a perfect indicator of whether or
not someone consumed tobacco. Dental calculus is also more transient
than for example bone, as it can be mechanically removed, intentionally
or unintentionally, during life, eliminating all trace of the alkaloids
consumed prior to its removal.\\
Quantitation of the detected compounds may have limited value in
archaeological samples due to degradation, and will greatly affect our
correlations related to concentration. Following burial, compound
stability over time will play a large role, as will microbial
degradation of compounds by bacteria and fungi in soil (Liu et al.,
2015), as well as the soil environment, such as temperature, pH, and
oxygen availability (Lindholst, 2010; Mackie et al., 2017).\\
The detected quantity of a compound will also depend on the quantity in
dental calculus during life, which is largely controlled by quantity of
consumption, how often the calculus was disrupted/removed, metabolic
breakdown of the compound, and inter- and intra-individual factors
related to stages of biofilm formation, maturation, and mineralisation
(Lustmann et al., 1976; Velsko et al., 2019; Zijnge et al., 2010). In
short, this means it is not really possible to detect the absence of a
compound. The absence of a compound is not evidence of absence of
consumption. This complicates the interpretation of our results. We have
attempted to minimise errors occurring due to this limitation by
including a relatively large sample of individuals and replicating our
analysis. Although given the relatively low detection rate seen in
tobacco, this remains a major limitation, and will likely be compounded
by increasing antiquity of the samples.

@@

All raw data is available on Zenodo
(\url{https://doi.org/10.5281/zenodo.7648757}). Analysis scripts, and
the source code for the manuscript and supplementary materials are
available as a research compendium
(\url{https://doi.org/10.5281/zenodo.7649825}) using the structure
recommended by the \textbf{rrtools} R package (Marwick, 2019).

@@

\leavevmode\vadjust pre{\hypertarget{ref-aten400JaarBeemster2012}{}}%
Aten, D., Bossaers, K. W. J. M., \& Misset, C. (2012). \emph{400 jaar
Beemster: 1612-2012}. {Stichting Uitgeverij Noord-Holland}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-leuwProhibitionLegalizationDutch1994}{}}%
Leuw, E., \& Marshall, I. H. (1994). \emph{Between {Prohibition} and
{Legalization}: {The Dutch Experiment} in {Drug Policy}}. {Kugler
Publications}. \url{https://books.google.com?id=2mAVkStNG5EC}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{lineno}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy1,; Jørgen B. Hasselstrøm2; Lambert K. Sørensen2; Maia Casna1; Menno Hoogland1; Historisch Genootschap Beemster3; Amanda G. Henry1},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 60\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, children and women.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, boxrule=0pt, sharp corners, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Hadley Wickham et al., 2019) and visualisations
were created using \textbf{ggplot2} (H. Wickham, 2016). Polychoric
correlations were calculated with the \textbf{psych} package (Revelle,
2022).

@@

\hypertarget{tbl-compound-detect}{}
\begin{longtable}[]{@{}lllr@{}}
\caption{\label{tbl-compound-detect}Target compound including whether it
was detected (TRUE) or not (FALSE) in each batch, as well as the lower
limit of quantitation (LLOQ) in ng. CBD = cannabidiol; CBN = cannabinol;
THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A;
THCVA = tetrahydrocannabivarin acid.}\tabularnewline
\toprule\noalign{}
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CBD & TRUE & FALSE & 0.050 \\
CBN & TRUE & FALSE & 0.050 \\
Caffeine & TRUE & TRUE & 0.050 \\
Cocaine & TRUE & FALSE & 0.025 \\
Cotinine & TRUE & TRUE & 0.050 \\
Nicotine & TRUE & TRUE & 0.100 \\
Salicylic acid & TRUE & TRUE & 0.500 \\
THC & TRUE & FALSE & 0.100 \\
THCA-A & TRUE & FALSE & 0.025 \\
THCVA & TRUE & FALSE & 0.010 \\
Theophylline & TRUE & TRUE & 0.010 \\
\end{longtable}

@@

\hypertarget{tbl-pearson}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0952}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}@{}}
\caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
dichotomous skeletal lesions and compound concentrations (ng/mg) from
the second batch. Correlations between pairs of dichotomous variables
are removed due to incompatibility with a Pearson correlation. OA =
osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD
= degenerative disc disease; CO = cribra orbitalia; CMS = chronic
maxillary sinusitis; SA = salicylic acid; PN = pipe
notches.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
OA & -0.12 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
VOP & -0.088 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
SN & -0.24 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
DDD & -0.0038 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 &
-0.0086 \\
CO & 0.064 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
CMS & -0.19 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
Caries & & -0.2 & -0.36 & -0.15 & -0.17 & -0.21 & -0.0045 & -0.22 \\
Nicotine & & & -0.21 & 0.01 & -0.014 & 0.43 & 0.14 & 0.98 \\
SA & & & & 0.14 & 0.37 & 0.038 & 0.17 & -0.17 \\
Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.031 \\
PN & & & & & & -0.16 & 0.18 & -0.0068 \\
Theophylline & & & & & & & 0.51 & 0.36 \\
Caffeine & & & & & & & & 0.078 \\
\end{longtable}

@@

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.847). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.487), SN and
cotinine (0.559), DDD and calculus (-0.416), CMS and caffeine (0.53),
caries and periodontitis (0.523), periodontitis and VOP (0.487),
periodontitis and age-at-death (0.407), nicotine and SN (0.53), calculus
and DDD (-0.416), age-at-death and theophylline (-0.45), theophylline
and age-at-death (-0.45), caffeine and periodontitis (0.494), cotinine
and CMS (0.427). Remaining correlations were weak or absent
(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

@@

Because we targeted individuals with moderate-to-large calculus
deposits, it is likely a biased sample. The presence of calculus may
increase the risk of premature death (Yaussy \& DeWitte, 2019), and
periodontal disease (which may or may not be associated with dental
calculus build-up) is a risk-factor for respiratory diseases, if
periodontal and respiratory pathogens enter the bloodstream (Azarpazhooh
\& Leake, 2006; Scannapieco, 1999; Scannapieco \& Ho, 2001). In our
sample, the percentage of chronic maxillary sinusitis (37.0\%) is lower
than in another (more representative) male sample (44.1\%) (Casna et
al., 2021), and the caries percentage is similarly lower in our sample
(17.6\%) than a more representative sample (22.9\%) (Lemmers et al.,
2013).\\
We used the presence/absence of a pipe notch and concurrent detection of
tobacco as a crude estimate of the accuracy of the method, which we
found to be around 59.3\%. This is a very rough estimate, as the
presence of a pipe notch is likely not a perfect indicator of whether or
not someone consumed tobacco. Dental calculus is also more transient
than for example bone, as it can be mechanically removed, intentionally
or unintentionally, during life, eliminating all trace of the alkaloids
consumed prior to its removal.\\
Quantitation of the detected compounds may have limited value in
archaeological samples due to degradation, and will greatly affect our
correlations related to concentration. Following burial, compound
stability over time will play a large role, as will microbial
degradation of compounds by bacteria and fungi in soil (Liu et al.,
2015), as well as the soil environment, such as temperature, pH, and
oxygen availability (Lindholst, 2010; Mackie et al., 2017).\\
The detected quantity of a compound will also depend on the quantity in
dental calculus during life, which is largely controlled by quantity of
consumption, how often the calculus was disrupted/removed, metabolic
breakdown of the compound, and inter- and intra-individual factors
related to stages of biofilm formation, maturation, and mineralisation
(Lustmann et al., 1976; Velsko et al., 2019; Zijnge et al., 2010). In
short, this means it is not really possible to detect the absence of a
compound. The absence of a compound is not evidence of absence of
consumption. This complicates the interpretation of our results. We have
attempted to minimise errors occurring due to this limitation by
including a relatively large sample of individuals and replicating our
analysis. Although given the relatively low detection rate seen in
tobacco, this remains a major limitation, and will likely be compounded
by increasing antiquity of the samples.

@@

All raw data is available on Zenodo
(\url{https://doi.org/10.5281/zenodo.8061483}). Analysis scripts, and
the source code for the manuscript and supplementary materials are
available as a research compendium
(\url{https://doi.org/10.5281/zenodo.7649825}) using the structure
recommended by the \textbf{rrtools} R package (Marwick, 2019).

@@

\leavevmode\vadjust pre{\hypertarget{ref-aten400Jaar2012}{}}%
Aten, D., Bossaers, K. W. J. M., \& Misset, C. (2012). \emph{400 jaar
Beemster: 1612-2012}. {Stichting Uitgeverij Noord-Holland}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-leuwProhibitionLegalization1994}{}}%
Leuw, E., \& Marshall, I. H. (1994). \emph{Between {Prohibition} and
{Legalization}: {The Dutch Experiment} in {Drug Policy}}. {Kugler
Publications}. \url{https://books.google.com?id=2mAVkStNG5EC}
